\pdftitle

### 2017-05-18 maxNt.tex
discussion

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-18T01:43:11+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

The notation in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. We intend \enquote{probability} as the degree of
belief which \enquote{would be agreed by all rational men if there were any
  rational men} \cite{good1966}.

@@

We need to make an initial probability assignment before any experimental
observations are made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental
observations.

We yet know very little about the physical details of the individual
neurons; their locations for example. Our initial state of knowledge $\yH$
is therefore symmetric, or \enquote{exchangeable}, under their
permutations. This symmetry must be reflected in our initial probability:
the \emph{representation theorem for finite exchangeability} states that it
must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \| \yH) = \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. Proof of
this theorem and generalizations to non-binary and continuum cases are
given by de~Finetti \cite{definetti1959b}, Ericson \cite{ericson1976},
Diaconis \cite{diaconis1977}, Heath \amp\ Sudderth \cite{heathetal1976}.
This theorem is intuitive: owing to symmetry, all states with $N\yRf$
active neurons must have equal probabilities.

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}

The latter conditional probability is a hypergeometric distribution
$\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without replacement}
problems. The combinatorial proof of the formulae above is in fact the same
as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

@@

\begin{figure}[!t]
\centering
%\includegraphics[width=0.75\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.75\linewidth]{pop_sample_projection2.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1} \defs
  \ypp(\yrs\|\yRf)$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, for
example \fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of
\enquote{fuzzy identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$
and $\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix.
We thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities; they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

Yet the approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. Upon observation
of a sample average, say $\yxxs=a$, the updated expectations for such
average in the population and in any new sample will usually be shifted
towards the observed value, as follows from Bayes's theorem and the
formulae above.

@@

The symmetry present in the probabilities above is not a physical property
of the neuronal population. It only expresses the symmetry of our initial
uncertainty about the population, and does not imply any sort of physical
similarity between the neurons. Subsequent observations may in fact break
this symmetry. The initial symmetry should intuitively also apply to the
sample, and this is indeed the case: the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

@@

\section{Enter maximum-entropy: quandary}
\label{sec:specific_initial_probability}

@@

The \me\ method is brought into play to solve this indeterminacy. It
selects one distribution, purported to be \enquote{maximally noncommittal},
among those that have the desired expectations. But we are in a quandary:
formulae~\eqref{eq:pullback_P} allow us to apply the method to find the
probability of the population $\p(\yXf=\yRf \| \yH)$, or of the sample
$\p(\yxs=\yrs \| \yH)$. \emph{The two applications, however, are
  inequivalent}. They lead to numerically different $\p(\yxs=\yrs \| \yH)$.

@@

The probabilities for the sample average obtained from application at the
population level~\eqref{eq:app_maxent_pop_marg} and at the sample level
\eqref{eq:app_maxent_sample} will be approximately equal, by our previous
observation~\eqref{eq:roughly_equal_nN} and also by the fact that they must
satisfy the same expectations for $\yf$. But the latter may miss important
features present in the former, like additional modes or particular tail
behaviour.

%\setlength{\intextsep}{0.5ex}%
%\begin{wrapfigure}{r}{0.5\linewidth}
\begin{figure}[!t]
\centering
\includegraphics[width=0.75\linewidth]{different_maxent_pop_sample_200_realdata_2mom.pdf}%
\caption{Plots of $\p(\yxs = \yrs)$ constructed by \me\ at the population
  level followed by sample marginalization (blue triangles),
  \eqn~\eqref{eq:full_pop_av_example_ME_marg}, and at the sample level (red
  circles), \eqn~\eqref{eq:sample_av_example_ME}. In this case $N=5000$,
  $n=200$, $\expe{\yxs}=0.0478$, $\expe{\yxxs}=0.00257$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb
% \end{wrapfigure}
\begin{figure}[!t]
\centering
\includegraphics[width=0.75\linewidth]{different_maxent_pop_sample_200_realdata_4mom.pdf}
% 
\caption{Linear and log-plots of $\p(\yxs = \yrs)$ constructed by \me\ at
  the population level followed by sample marginalization (blue triangles),
  \eqn~\eqref{eq:full_pop_av_example_ME_marg}, and at the sample level (red
  circles), \eqn~\eqref{eq:sample_av_example_ME}. In this case $N=5000$,
  $n=200$, $\expe{\yxs}=0.0478$, $\expe{\yxxs}=0.00257$,
  $\expe{\sav{\yx\yx\yx}}=1.48\times10^{-4}$,
  $\expe{\sav{\yx\yx\yx\yx}}=8.81\times 10^{-6}$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
We show two examples of this discrepancy in
\figs~\ref{fig:diff_maxent_pop_sample}
and~\ref{fig:diff_maxent_pop_sample_realdata}, for $N=5000$, $n=200$. In
the first example the constraints are $\expe{\yxs} = c_1$ and
$\expe{\yxxs} = c_2$, with $c_1=0.0478$ and $c_2=0.00257$. The distribution
obtained at the sample level is broader than the one obtained at the
population level; the tails of the two distributions are very different.
The second example includes two additional constraints
$\expe{\sav{\yx\yx\yx}}= c_3$, $\expe{\sav{\yx\yx\yx\yx}}=c_4$ with
$c_3=0.000148$, $c_4=8.81\times 10^{-6}$. The distribution obtained at the
population level has two modes, replaced by only one in the distribution
obtained at the sample level; the tails are very different also in this
case. The constraints used in these examples have neurobiologically
realistic values \cite{rostamietal2016}.

The inequivalence of the two \me\ applications is generally valid for small
number of constraints, even if the two states of knowledge $\yHa$, $\yHb$
use different numbers $m'$, $m''$ \emph{and values} of constraints. The
equality $\p(\yxs = \yrs \| \yHa) = \p(\yxs = \yrs \| \yHb)$ in this case
requires solving $n$ equations in $m'+m''$ unknowns (normalization
is taken care of), so in general it can only be satisfied if $n\le m'+m''$.

@@

Where should we apply the \me\ method then, on the sample or on the
population?

@@

The question that closed the preceding section cannot receive a categorical
answer. An answer can only be given case by case, depending on the
inferences we are trying to make, on the assumptions needed and those
negligible, and on our computational power. We want to stress, in fact, the
extreme importance and priority of the following two questions:
\begin{itemize}
\item Of which quantities do we need to quantify the uncertainty? and why?
\item What are the assumptions we need or want to make to answer the
  question above?
\end{itemize}

The first purpose of this note was to remind our readers that probability
models can have hidden assumptions that we don't want to make or that are
in contradiction with assumptions we want to make. The dilemma generated by
the \me\ method and the assumption of random representative sampling is an
example. The usual application of \me\ at the sample level generates a
different distribution than the application at the population level. Then
the former application must implicitly assume that no larger population
exist; or in other words that the neurons of the sample are isolated from
their environment.

The probability distribution for the sample average obtained by applying
\me\ at the population level has by construction a lower entropy than the
one obtained by application at the sample level. The former distribution
says that some sample states $\yx$ should be considered less likely than in
the latter distribution. A lower entropy signals the presence of an
additional assumption. Such assumption, intuitively, is that the sample is
part of a larger population, and our ignorance about the former is
representative of our ignorance about the latter. This assumption may
affect our predictions.

In a neuroscientific context we find this assumption very natural, and
therefore the \me\ method at the population level preferable. Also
\emph{physical} neuronal-network models add indeed an external input to the
neurons, mimicking their embedding in a larger network.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-18T11:35:39+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

The notation in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. By \enquote{probability} we mean plausibility or
the degree of belief which \enquote{would be agreed by all rational men if
  there were any rational men} \cite{good1966}.

@@

We need to make an initial probability assignment before any experimental
observations are
made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental observations.
We would also like our probability assignment to reflect that the sample is
somehow \enquote{representative} of the population.

Let's assume that we initially know very little about the physical details
of the individual neurons; their locations for example. Our initial state
of knowledge or ignorance $\yH$ is therefore symmetric, or
\enquote{exchangeable}, under their permutations. This symmetry must be
reflected in our initial probability: the \emph{representation theorem for
  finite exchangeability} states that it must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \| \yH) = \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. Proof of
this theorem and generalizations to non-binary and continuum cases are
given by de~Finetti \cite{definetti1959b}, Ericson \cite{ericson1976},
Diaconis \cite{diaconis1977}, Heath \amp\ Sudderth \cite{heathetal1976}.
This theorem is intuitive: owing to symmetry, we must assign equal
probabilities to all states with $N\yRf$ active neurons.

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
Our initial symmetric ignorance should intuitively also apply to the
sample; indeed, the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well. The conditional probability
in the last formula is a hypergeometric distribution $\ypp(\yrs\|\yRf)$,
typical of \enquote{drawing without replacement} problems. The
combinatorial proof of the formulae above is in fact the same as for this
class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

@@

\begin{figure}[!t]
\centering
%\includegraphics[width=0.75\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.75\linewidth]{pop_sample_projection2.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
\ypp(\yrs\|\yRf) \defd  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, see
\fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of \enquote{fuzzy
  identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$ and
$\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix. We
thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities: they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

Yet, the approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. This is how the
idea of representativeness is translated into our probabilities. The
symmetry present in our initial probabilities is not a physical property of
the neuronal population or sample. It only expresses the symmetry of our
initial uncertainty about them, and does not imply any sort of physical
similarity between the neurons. Subsequent observations may in fact break
this symmetry. Upon observation of a sample average, say $\yxxs=a$, the
updated expectations for such average in the population and in any new
sample will usually be shifted towards the observed value, as follows from
Bayes's theorem and the formulae above.

@@

@@

\section{Enter maximum-entropy: dilemma}
\label{sec:specific_initial_probability}

@@

The \me\ method is brought into play to solve this indeterminacy. It
selects one distribution, purported to be \enquote{maximally noncommittal},
among those that have the desired expectations. But here's a dilemma:
formulae~\eqref{eq:pullback_P} allow us to apply the method to find the
probability of the population $\p(\yXf=\yRf \| \yH)$, or of the sample
$\p(\yxs=\yrs \| \yH)$. \emph{The two applications, however, are
  inequivalent}. They lead to numerically different $\p(\yxs=\yrs \| \yH)$.

@@

The probabilities for the sample average obtained from application at the
population level~\eqref{eq:app_maxent_pop_marg} and at the sample level
\eqref{eq:app_maxent_sample} should be approximately equal, by our previous
observation about representativity~\eqref{eq:roughly_equal_nN} and also by
the fact that they must satisfy the same expectations for $\yf$.

Yet they cannot be exactly equal, because their equality would require the
Lagrange multipliers $\yL$ and $\yl$ to satisfy equations
\eqref{eq:app_maxent_pop_constraints},
\eqref{eq:app_maxent_sample_constraints}, and
$\p(\yxs = \yrs \| \yHa) = \p(\yxs = \yrs \| \yHb)$ -- that is, $2m+n$
equations in $m$ unknowns. This could be possible, if at all, only for very
special choices of constraints functions $\yf$ and values $\yc$.

The sample distribution obtained from \me\ at the sample level will
therefore likely miss important features present in the one obtained at the
population level, like additional modes or particular tail behaviour.

%\setlength{\intextsep}{0.5ex}%
%\begin{wrapfigure}{r}{0.5\linewidth}
\begin{figure}[!t]
\centering
\includegraphics[width=0.99\linewidth]{different_maxent_pop_sample_200_realdata_2mom.pdf}%
\caption{Linear and log-plots of $\p(\yxs = \yrs)$ constructed by \me\ at the population
  level followed by sample marginalization (blue triangles),
  \eqn~\eqref{eq:full_pop_av_example_ME_marg}, and at the sample level (red
  circles), \eqn~\eqref{eq:sample_av_example_ME}, with $N=5000$,
  $n=200$, constraints $\expe{\yxs}=0.0478$, $\expe{\yxxs}=0.00257$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb
% \end{wrapfigure}
\begin{figure}[!t]
\centering
\includegraphics[width=0.99\linewidth]{different_maxent_pop_sample_200_realdata_4mom.pdf}
% 
\caption{Linear and log-plots of $\p(\yxs = \yrs)$ constructed by \me\ at
  the population level followed by sample marginalization (blue triangles),
  \eqn~\eqref{eq:full_pop_av_example_ME_marg}, and at the sample level (red
  circles), \eqn~\eqref{eq:sample_av_example_ME}, with $N=5000$,
  $n=200$, constraints $\expe{\yxs}=0.0478$, $\expe{\yxxs}=0.00257$,
  $\expe{\sav{\yx\yx\yx}}=1.48\times10^{-4}$,
  $\expe{\sav{\yx\yx\yx\yx}}=8.81\times 10^{-6}$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
We show two examples of this discrepancy in
\figs~\ref{fig:diff_maxent_pop_sample}
and~\ref{fig:diff_maxent_pop_sample_realdata}, for $N=5000$, $n=200$. In
the first example the constraints are $\expe{\yxs} = c_1$ and
$\expe{\yxxs} = c_2$, with $c_1=0.0478$ and $c_2=0.00257$. The distribution
obtained at the sample level is broader than the one obtained at the
population level; the tails of the two distributions are very different.
The second example includes two additional constraints
$\expe{\sav{\yx\yx\yx}}= c_3$, $\expe{\sav{\yx\yx\yx\yx}}=c_4$ with
$c_3=0.000148$, $c_4=8.81\times 10^{-6}$. The distribution obtained at the
population level has two modes, replaced by only one in the distribution
obtained at the sample level; the tails are very different also in this
case. The constraints used in these examples have neurobiologically
realistic values \cite{rostamietal2016}.

@@

How should we apply the \me\ method then? on the sample or on the
population? Which one is maximally noncommittal?

@@

The question that closed the preceding section cannot receive a categorical
answer. An optimal answer can only be given case by case, depending on the
computational power available, on which inferences we are trying to make,
on which assumptions we need or want to make and those we wish to avoid.

\iffalse We want to stress, in fact, the
extreme importance and priority of the following two questions:
\begin{itemize}
\item Of which quantities do we need to quantify the uncertainty? and why?
\item What are the assumptions we need or want to make to answer the
  question above?
\end{itemize}\fi

The first purpose of this note was indeed to remind ourselves that
probability models require a continuous scrutiny, because they can rest on
hidden assumptions that we don't want to make or that contradict others we
are making. The dilemma generated by the \me\ method and the assumption of
random representative sampling is an example. The usual application of \me\
at the sample level yields a different distribution from the application at
the population level. The former application must therefore implicitly
assume that no larger population exist, or that the neurons of the sample
are isolated from their environment.

The \me\ application at the population level has in fact by construction a
lower entropy than the one obtained by application at the sample level:
both satisfy the same constraints, but only the latter is obtained by
direct maximization. The lower entropy of the distribution obtained from
application at the population level is clearly visible from its higher
sharpness in \fig~\ref{fig:diff_maxent_pop_sample}. This distribution
therefore says that some sample states $\yx$ should be considered less
likely than in the distribution obtained at the sample level. A lower
entropy signals the presence of additional assumptions or constraints. In
this case the assumption is that the sample is part of a larger population,
and our ignorance about the sample should be representative of our
ignorance about a larger population. This assumption affects our
predictions.

In a neuroscientific context we find this assumption very natural, and
therefore the \me\ method at the population level preferable. After all,
also \emph{physical} neuronal-network models usually include some sort of
external input to the neurons, mimicking their embedding in a larger
network.
